---
title: "cancer_blood_biomarkers"
author: "Hansen Han"
date: "2024-06-18"
output: html_document
---

# Cancer Blood Biomarker Discovery (Analysis)
Analysis to identify diagnostic biomarkers of cancer via meta-analysis of gene expression studies.

## Load Libraries
```{r messages=FALSE errors=FALSE}
library(MetaIntegrator)
library(tidyverse)
library(magrittr)
library(ggpubr)
library(biomaRt)
```

## Load Processed Data
```{r}
load("processed_cancer_blood_datasets.Rdata")
```


## Split into Test & Train

Split 
```{r}
# Set seed for reproducibility
set.seed(123)

study_ids <- c(
  "GSE198048",
  "GSE138118",
  "GSE136651",
  "GSE74629",
  "GSE20189",
  "GSE39345",
  "GSE120691",
  "GSE164191",
  "GSE49515",
  "GSE27562",
  "GSE10715"
)

# Shuffle the list
shuffled_ids <- sample(study_ids)

# Split into train and test sets
train_ids <- shuffled_ids[1:6]
test_ids <- shuffled_ids[7:11]
test_ids <- c(test_ids, "GSE16443") # add this one in there no matter what since it only has 6k probes

# Print the results
cat("Train IDs:\n")
print(train_ids)
cat("\nTest IDs:\n")
print(test_ids)
```

Format datasets
```{r}
# remove train datasets from test data and vice versa
test_data <- blood_biomarker_data_raw
for (study_to_delete in train_ids) {
  test_data$originalData[[study_to_delete]] <- NULL
}

train_data <- blood_biomarker_data_raw
for (study_to_delete in test_ids) {
  train_data$originalData[[study_to_delete]] <- NULL
}

```

## Meta Analysis

### Calculate Effect Sizes
```{r}
meta_results <- runMetaAnalysis(train_data, runLeaveOneOutAnalysis= TRUE, maxCores=Inf)
```

### Identify Gene Signature
Here, we look for a gene signature with an effect size > 1, significance of =< 0.05
```{r}
meta_results <- filterGenes(meta_results, isLeaveOneOut = FALSE, effectSizeThresh = 1, FDRThresh = 0.05)
```

View gene signature
```{r}
meta_results$filterResults[[1]]
```

```{r}
heatmapPlot(metaObject = meta_results, filterObject = meta_results$filterResults[[1]])
```


### Test Accuracy

Performance on discovery data is varied but overall 0.87 AUC
```{r}
summaryROCPlot(metaObject = meta_results, filterObject = meta_results$filterResults[[1]], bootstrapReps = 100)
```

Performance on validation data breaks down with a summary AUC of 0.55
```{r}
summaryROCPlot(metaObject = test_data, filterObject = meta_results$filterResults[[1]], bootstrapReps = 100)
```

